The Private School Network: Recruiting Visits to Private High Schools by Public and Private Universities

### 2023-09-11 r_r/front_end.Rmd
private hs data

\newpage

@@

PX-PX: vignette of visits to a private high school.

Theories of organizational behavior suggest that university recruiting behaviors -- those that require significant resource expenditure -- are indicators of organizational enrollment priorities. Large, complex organizations consist of several "technical levels" responsible for processing inputs and producing outputs and "managerial levels" responsible for directing the technical level and tending to demands from the external environment [@RN531; @RN959]. The technical level cannot pursue multiple goals efficiently. Thus, organizations respond to external pressue to pursue multiple, competing goals goals by symbolically adopting some goals and substantively adopting others [@RN513]. Under symbolic adoption, organizations adopt policies and structures that send signals to external constituents but do not affect resource allocation in the technical level. Under substantive adoption, organizations direct resources in the technical level towards achieving the goal. Off-campus recruiting visits by admissions officers to local high schools and college fairs was the top-ranked recruiting expenditure item for both public and private universities, according to @RN4664. Therefore, analyses of which schools a university visits -- and which schools they ignore -- can yield insights about university enrollment goals.

@@

A multidisciplinary literature on social stratification has analyzed relationship between private high schools and universities [CITE]. This research observes strong linkages -- and flows of students -- between private high schools and selective private universities [CITE]. To date, scholarship on nonresident enrollment by public universities has progressed in isolation from scholarship on linkages between private schools and private universities. We argue that these two stories are connected because public research universities are increasingly targeting out-of-state private schools, which have historically been the domain of selective private universities.

<!-- 
This study compares recruiting visits to private high schools by public research universities and recruiting visits to private high schools by selective private universities. 
We argue that recruiting visits are an indicator of enrollment goals -- the university wants to enroll some students from the school, the school wants some students to enroll at the university. 
-->
This study analyzes recruiting visits to private high schools made in the 2017 calendar year by a sample of 15 public research universities and 14 selective private universities. @RN3519 finds that travel to local high schools is vital for maintaining relationships with guidance counselors at "feeder schools." We conceptualize recruiting visits as an indication the university and the school have a social relationship. This conceptualization motivates the use of social network methods, which analyze networks defined by "network ties" (i.e., visits) between actors (universities and schools). Comparing the recruiting networks of public researc universities and selective private universities yields novel insights about competition for students at schools that enroll affluent, "full-pay" students. Analyses are informed by three research questions:

>>>>

\pagenumbering{roman}
\newpage
\pagenumbering{gobble}
\pagenumbering{arabic}

@@

Autumn is college visit season at Harpeth Hall, a non-sectarian private school for girls in Nashville, TN, founded in 1951. According to the 2020 *Best Private High Schools in America* by Niche, Harpeth is an "A+" school ranked 337 in the nation. Enrollment growth at Harpeth and the Nashville-area private school market mirror broader regional patterns, whereby “efforts by the courts to desegregate public schools in the South resulted in White flight to private schools” [@RN4504, p. 14]. In Fall 2017, about 85% of Harpeth students were white and less than 5% were Black or Latinx (Author calculations).

During the 2017 calendar year, we tracked off-campus recruiting visits made by 14 selective private universities, 12 selective private liberal arts colleges, and 15 public research universities.^[NAME THE UNIVERSITIES]. We are still trying to figure out what happened.

From September 6th through October 17th, Harpeth received recruiting visits from 11 private universities (Northwestern, Boston College, Tufts, Villanova, Case Western Reserve, University of Denvor, Emory, Tulane, Baylor, Texas Christian University), and three private liberal arts colleges from our sample (Harvey Mudd, Scrips, and Sewanee, which visited twice). Harpeth also received visits from the University of Cincinatti, University of South Carolina, the University of Alabama, University of Arkansa, University of Georgia, and CU Boulder.

Four miles northeast of Harpeth, next to Vanderbilt University, is the University School of Nashville. The University School is a nonsectarian, "A+" school, ranked 127th in the nation, and is more racially diverse than Harpeth (64% white, 11.5% Black, 3.5% Latinx, 10.0% Asian). From September 11th 2017 through October 18th, The University School received visits from the same 11 private universities that visited Harpeth (except Notre Dame visited and Baylor did not), five liberal arts colleges (Williams, Middlebury, Sewaneee, Scripps, and Harvey Mudd), but only two public research universities (University of Cincinatti, and CU Boulder). Four miles south of The University School lies Father Ryan High School, a predominantly white (86%, 7% Black or Latinx) Catholic school that received an "A" grade by Niche. In fall 2017, Father Ryan received visits from nine of the 14 private universities in our sample (including all the Catholic universities), just one liberal arts college (Sewanee), and the same six public universities that visited Harpeth Hall, including two visits by the University of Alabama.

Theories of organizational behavior suggest that university recruiting behaviors -- those that require significant expenditure -- are indicators of organizational enrollment priorities. Large, public-facing organizations are beset by multiple, competing pressures from the external environment, but organizations have difficulty pursuing multiple goals [@RN531; @RN959]. Thus, organizations respond by symbolically adopting some goals and substantively adopting others [@RN513]. Under symbolic adoption, organizations adopt policies and structures that send signals to external constituents but do not substantively affect resource allocation inside the the "technical level" of the organization.^[Contingency theory states that organizations consist of "technical levels" responsible for processing inputs and producing outputs and "managerial levels" responsible for directing the technical level and tending to demands from the external environment [@RN531]. The technical level achieves effectiveness by focusing on a small, stable set of goals.] Under substantive adoption, organizations direct resources in the technical level towards achieving the goal. Off-campus recruiting visits to local high schools and college fairs was the top-ranked recruiting expenditure item for both public and private universities, according to @RN4664. Therefore, analyses of which schools a university visits -- and which schools they ignore -- can yield insights about university enrollment goals.

@@

Within scholarship on social mobility and social stratification, a multidisciplinary literature analyzes the relationship between private school and college [CITE]. This research observes strong links -- and student flows -- between private high schools and selective private universities [CITE]. To date, scholarship on nonresident enrollment by public universities has progressed in isolation from scholarship on links between between private school and college. We argue that these two stories have become connected because public research universities increasingly target out-of-state private schools, schools that have been the domain of selective private universities.

<!-- 
This study compares recruiting visits to private high schools by public research universities and recruiting visits to private high schools by selective private universities. 
We argue that recruiting visits are an indicator of enrollment goals -- the university wants to enroll some students from the school, the school wants some students to enroll at the university. 
-->
This study analyzes recruiting visits to private high schools made in the 2017 calendar year by a sample of 15 public research universities and 14 selective private universities. @RN3519 finds that travel to local high schools is vital for maintaining relationships with guidance counselors at "feeder schools." We conceptualize recruiting visits as an indication the university and the school have a social relationship. This conceptualization motivates the use of social network methods, which analyze networks defined by "network ties" (i.e., visits) between actors (universities and schools). Comparing the recruiting networks of public research universities and selective private universities yields novel insights about competition for students at schools that enroll affluent, "full-pay" students. Analyses are informed by three research questions:
